Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-02-10 rnni_convexity.tex
quick look at partition lattices

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to show that $\rnni$ does not have the Split Property and consider a variant, the Weak Split Property, which says that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well known algebraic structure.

@@

In this section we establish a connection between the $\rnni$ graph and a well known algebraic structure, the partition lattice.
The relation between these two provides a new direction for further research and leads to a different view on problems of either structure.

The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n,\leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$.
For simplification we will denote the partition lattice by $\Pi_n$, $\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
This algebraic structure is related to the $\rnni$ graph in the following way.

>>>>

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
Splits are denoted by $A|B$.
We will give a counterexample to show that $\rnni$ does not have the Split Property and consider a variant, the Weak Split Property, which says that every shared split of taxa is maintained along a shortest path.
We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well-known algebraic structure.

@@

In this section we establish a connection between the $\rnni$ graph and a well-known algebraic structure, the partition lattice.
This connection provides a new direction for further research and translates results and open problems from the language of phylogenetics to the language of lattice theory.

The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$, which means...
For simplification we will denote the partition lattice on $n$ elements by $\Pi_n$.
$\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
This algebraic structure is related to the $\rnni$ graph in the following way.
